---
title: "Reproducible_Research_Course_Project_1"
author: "Luis OLiveira"
date: "15/08/2020"
output: word_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

### Loading Data
The data file "activity.csv" already unziped should be in the user current R working directory.

```{r} 
MainDataAtivity <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric")) 
```

## ----What is mean total number of steps taken per day?---

## Exclude NA steps rows
```{r} 
names(MainDataAtivity) 
str(MainDataAtivity) 
head(MainDataAtivity[which(!is.na(MainDataAtivity$steps)), ]) 
```

## Compute the total steps per day
```{r} 
totalstepsday <- aggregate(steps ~ date, MainDataAtivity, sum) 
head(totalstepsday) 
```

## How to build the Histogram of the total number of steps taken each day
```{r} 
paletteBlue <- colorRampPalette(c("skyblue", "darkblue", "skyblue")) 
hist(totalstepsday$steps, breaks=10, xlab="Number of steps",  ylab="Number of days",
main="Histogram - Total number of steps per Day", 
col=paletteBlue(22), family="serif") 
```

## Computing the mean and median number of steps taken each day
```{r} 
library(dplyr) 
stepsstatistics <- summarise(totalstepsday, meantotalsteps=mean(totalstepsday$steps),                              
mediantotalsteps=median(totalstepsday$steps)) 
print(stepsstatistics) 
```

## -----What is the average daily activity pattern?----
```{r}
## Time series plot of the average number of steps taken

meanstepsinterval <- aggregate(steps ~ interval, MainDataAtivity, mean) 
head(meanstepsinterval) 

plot(x=meanstepsinterval$interval, y=meanstepsinterval$steps, type="l", 
main="Time Series Plot - Average steps by per Interval", 
ylab="Number of steps", xlab="Intervals (in 5 minutes)", 
col="darkblue", lwd=1.5, family="serif")
```

## The 5-minute interval that, on average, contains the max number of steps
```{r}
meanstepsinterval[grep(max(meanstepsinterval$steps), meanstepsinterval$steps), ]
```

## -----Imputing missing values -----
## Code for imputing missing data
## Replace NA values in the steps variable by the mean of the interval

```{r}
adjusteddata <- MainDataAtivity 
for(x in 1:17568) { 
   if(is.na(adjusteddata[x, 1])==TRUE) { adjusteddata[x, 1] <- meanstepsinterval[meanstepsinterval$interval %in% adjusteddata[x, 3], 2] } } 
head(adjusteddata) 
```


## How to build the Histogram of the total number of steps taken each day after missing values are imputed

```{r}
adjustedtotalstepsday <- aggregate(steps ~ date, adjusteddata, sum) 
head(adjustedtotalstepsday) 

paletteRed <- colorRampPalette(c("deeppink", "darkred", "deeppink")) 
hist(adjustedtotalstepsday$steps, breaks=20, xlab="Number of Steps Taken",  ylab="Number of Days",
main="Histogram - Total Number of Steps per Day (Adjust. values)", 
col=paletteRed(22), family="serif") 
```


## How to compare Histograms to see the impact of imputing vakues in missing data

```{r}
par(mfrow = c(1, 2)) 
hist(totalstepsday$steps, breaks=20, xlab="Number of Steps Taken",  ylab="Number of Days",
col=paletteBlue(22), family="serif", ylim=c(0, 20), main=NULL) 
hist(adjustedtotalstepsday$steps, breaks=20, xlab="Number of Steps Taken Adjusted",  ylab="Number of Days",
col=paletteRed(22), family="serif", ylim=c(0, 20), main=NULL) 
mtext("Histograms -Total Number of Steps Taken per Day, Original vs Adjusted Data", 
adj=0.95, family="serif", font=2) 
```

##Analysis if there are differences in activity patterns between weekdays and weekends?

```{r}
daysdata <- adjusteddata 
daysdata$days <- weekdays(daysdata$date) 
daysdata$weekday <- as.character(rep(0, times=17568)) 
for(x in 1:17568) { 
    if(daysdata[x, 4] %in% c("Saturday", "Sunday")) { daysdata[x, 5] <- "Weekend" } else { daysdata[x, 5] <- "Weekday" } } 
daysdata$weekday <- factor(daysdata$weekday) 
head(daysdata) 
```

## Comparing the weekday and weekend data

```{r}
weekdaydata <- daysdata[daysdata$weekday=="Weekday", ] 
weekenddata <- daysdata[daysdata$weekday=="Weekend", ] 

weekdayMean <- aggregate(steps ~ interval, weekdaydata, mean) 
weekendMean <- aggregate(steps ~ interval, weekenddata, mean) 
```

## Making the plot chart to compare the two series
```{r}
par(mar = rep(2, 4))
plot(weekdayMean$interval, weekdayMean$steps, type="l", 
main="Avg. Steps per Interval - Weekdays", 
xlab="Intervals (5 mins)", ylab="Number Steps", family="serif", 
col="darkred", lwd=1.5, ylim=c(0, 230)) 
plot(weekendMean$interval, weekendMean$steps, type="l", 
main="Avg. Steps per Interval - Weekends", 
xlab="Intervals (5 mins)", ylab="Number Steps", family="serif", 
col="darkblue", lwd=1.5, ylim=c(0, 230)) 
```
